---
title: "Reticulate"
subtitle: "RaukR 2021 • Advanced R for Bioinformatics"
author: "<b>Nina Norgren</b>"
institute: NBIS, SciLifeLab
keywords: r, RaukR, markdown
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/presentation.css'
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "RaukR 2021 • %current%/%total%"
---
exclude: true
count: false

```{r,echo=FALSE,child="assets/header-presentation.Rmd"}
```

<!-- ----------------- Only edit title & author above this ----------------- -->

```{r,echo=FALSE,message=FALSE,warning=FALSE}
# load the packages you need

library(reticulate)
use_condaenv("raukr", required = TRUE)
library(tidyverse)
```
---
name: contents

## Contents

* [Learning outcomes](#LOs)
* [Introduction](#intro)
* [Introducing reticulate](#reticulate)
* [Importing modules](#importing_modules)
* [Sourcing scripts](#sourcing_scripts)
* [Executing Python code](#execute_code)
* [Python in R Markdown](#python_rmarkdown)
* [Type conversions](#type_conversions)
* [Examples](#examples)

---
name: LOs

## Learning outcomes

<br>

In this session we will learn to:

* Understand the concepts needed for running Python in R
* Understand the different object classes in Python and their equivalent in R
* Apply our knowledge to:
  + Import Python functions into R
  + Use R objects as input to Python functions
  + Translate between Python and R objects


---
name: intro

## Introduction

<br><br>
.largest.center[**R versus Python**]
###.center[The ultimate fight!]
--
name: intro2

<br><br><br>
###.center[Not anymore!] 



---
name:reticulate

## Introducing reticulate

* Combine Python and R code
* Use R classes in Python functions and vice versa
* Import Python functions into R code and run from R
* Add Python code chunks to markdown documents

```{r, eval= FALSE}
library(reticulate)
```

---
name:importing_modules

## Importing Python modules

```{r}
datetime <- import("datetime")
todays_r_date <- datetime$datetime$now()
```

--
```{r}
todays_r_date
class(todays_r_date)
```

--
Objects are automatically converted to R types, unless otherwise specified

--
```{r}
datetime <- import("datetime", convert = FALSE)
todays_py_date <- datetime$datetime$now()
```

--
```{r}
todays_py_date
class(todays_py_date)
```

---
## Importing built-in Python functions

Access Python's built-in functions directly in R

```{r}
builtins <- import_builtins()
r_vec <- c(1, 5, 3, 4, 2, 2, 3, 2)
str(r_vec)
```
r_vec is an R object.

--
```{r}
builtins$len(r_vec)
builtins$max(r_vec)
```
Python built-in functions still working on R objects

--
```{r}
max(r_vec)
```
Normal R way

---
name: sourcing_scripts

## Sourcing scripts

Import your own python functions for use in R. File `python_functions.py`:

```{python, evaluate=FALSE}
def add(x, y):
  return x + y
```

--
R code:

```{r}
source_python("python_functions.py")
class(4)
res <- add(4,5)
res
class(res)
```
--
Type `numeric` in and type `numeric` out. But what happens in between?

--
File `python_functions.py`:
```{python}
def add_with_print(x, y):
  print(x, 'is of the python type ', type(x))
  return x + y
```

```{r}
res2 <- add_with_print(4,5)
py_capture_output(add_with_print(4,5))
str(res2)
```

---
name: execute_code

## Execute Python code


Run python string:
```{r}
py_run_string("result = [1,2,3]*2")
py$result
```
All objects created by python are accessible using the `py` object exported by reticulate

--

Run python script `my_python_script.py`:

```{python}
def add(x, y):
  return x + y

def multiply_by_3(x):
  return x*3

def run_all():
  x = 5
  y = 8
  added = add(x, y)
  final = multiply_by_3(added)
  return final

final = run_all()
```

```{r}
py_run_file("my_python_script.py")
py$final
```

---
name: python_rmarkdown

## Python in R Markdown

In R Markdown it is possible to mix in Python chunks:

```{r,echo=FALSE,comment=""}
cat(paste0("```{python}\n","import pandas as pd\n\n","movies = get_all_movies()\n",
           "print(type(movies))\n", "``` \n"))
```


```{python, echo=FALSE}
import pandas as pd
from imdb_functions_pres import *

movies = get_all_movies(fromYear = 1960)
print(type(movies))
```

---
## Python in R Markdown

Access the movie object using the `py` object, which will convert movies to an R object:

```{r, eval = FALSE}
movies_r <- py$movies
movies_r <- as_tibble(movies_r)
subset <- movies_r %>% select(5:6, 8:10)
```

---
## Python in R Markdown

Access the movie object using the `py` object, which will convert movies to an R object:

```{r}
movies_r <- py$movies
movies_r <- as_tibble(movies_r)
subset <- movies_r %>% select(5:6, 8:10)
knitr::kable(subset[1:7,],'html')
```
---
## Python in R Markdown

Continue working with the now converted R object in R

```{r, eval = FALSE, tidy = FALSE}
ggplot(movies_r, aes(x=startYear)) + geom_bar() + 
                                     theme(axis.text.x = element_text(angle = 90)) +
                                     ggtitle('Number of movies per year')
```

---
## Python in R Markdown

Continue working with the now converted R object in R

```{r, fig.height = 4, fig.width = 8, tidy = FALSE}
ggplot(movies_r, aes(x=startYear)) + geom_bar() + 
                                     theme(axis.text.x = element_text(angle = 90)) +
                                     ggtitle('Number of movies per year')
```

---
name: type_conversions

## Type conversions

When calling python code from R, R data types are converted to Python types, and vice versa, when values are returned from Python to R they are converted back to R types.

####.center[**Conversion table**]

```{python, echo=FALSE}
import pandas as pd
d = {'R': ['Single-element vector', 'Multi-element vector', 'List of multiple types',
           'Named list', 'Matrix/Array', 'Data Frame', 'Function', 'Raw', 'NULL, TRUE, FALSE'],
     'Python': ['Scalar', 'List', 'Tuple', 'Dict', 'NumPy ndarray', 'Pandas DataFrame',
                'Python function', 'Python bytearray', 'None, True, False'],
     'Examples': [['1', '1L', 'TRUE', "foo"], ['c(1.0, 2.0, 3.0)', 'c(1L, 2L, 3L)'], ['list(1L, TRUE, "foo")'], 
                  ['list(a = 1L, b = 2.0)', 'dict(x = x_data)'], ['matrix(c(1,2,3,4), nrow=2, ncol=2)'], 
                  ['data.frame(x = c(1,2,3), y = c("a","b","c"))'], ['function(x) x +1'],
                  ['as.raw(c(1:10))'], ['NULL, TRUE, FALSE']]            
                }
df = pd.DataFrame(data=d)
```
```{r, echo=FALSE}
knitr::kable(py$df,'html')
```
---
## Type conversions

`python_functions.py`:

```{python}
def check_python_type(x):
  print(type(x))
  return x
```

--
```{r}
source_python("python_functions.py")

r_var <- matrix(c(1,2,3,4),nrow=2, ncol=2)
class(r_var)
py_capture_output(check_python_type(r_var))
r_var2 <- check_python_type(r_var)
class(r_var2)
```

--
```{r}
source_python("python_functions.py", convert=FALSE)

r_var <- matrix(c(1,2,3,4),nrow=2, ncol=2)
class(r_var)
py_capture_output(check_python_type(r_var))
r_var2 <- check_python_type(r_var)
class(r_var2)
r_var3 <- py_to_r(r_var2)
class(r_var3)
```
---
## Type conversions

- `42` in R is a floating point number. In Python it is an integer
```{r}
str(42)
check_python_type(42)
py_capture_output(check_python_type(42))
```
--
```{r}
str(42L)
check_python_type(42L)
py_capture_output(check_python_type(42L))
```

---
## Type conversions

- List conversions of single element vectors, automatically translated to Python scalar
```{r}
str(c(24))
check_python_type(c(24))
py_capture_output(check_python_type(c(24)))
```
--
```{r}
str(list(24))
check_python_type(list(24))
py_capture_output(check_python_type(list(24)))
```

---
name: examples

## Examples

<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end-slide
class: end-slide, middle
count: false

# Thank you. Questions?

```{r,echo=FALSE,child="assets/footer-presentation.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
rmarkdown::render("presentation.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("presentation.html",output="presentation.pdf")
```
