---
title: "Vectorization in R"
subtitle: "RaukR 2021 • Advanced R for Bioinformatics"
author: "<b>Marcin Kierczak</b>"
institute: NBIS, SciLifeLab
keywords: r, RaukR, markdown
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/presentation.css'
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "RaukR 2021 • %current%/%total%"
---
exclude: true
count: false

```{r,echo=FALSE,child="assets/header-presentation.Rmd"}
```

<!-- ----------------- Only edit title & author above this ----------------- -->

```{r,echo=FALSE,message=FALSE,warning=FALSE}
# load the packages you need

library(tidyverse)
library(peakRAM)
#library(ggplot2)
```
---
name: contents

## Learning Outcomes

By the end of this module, you will:

* understand how to write more efficient loops
* be able to vectorize most loops
* understand how the `apply*` functions work
* be aware of the `purrr` package
* understand what a recursive call is

---
name: for_loop_example
## The simplest of all `for` loops

Say, we want to add 1 to every element of a vector:

```{r for.loop.ex1, echo=T}
vec <- c(1:5)
vec
for (i in vec) {
  vec[i] <- vec[i] + 1
}
vec
```

--

Exactly the same can be achieved in R by means of **vectorization**:
```{r for.loop.avoid, echo=T}
vec <- c(1:5)
vec + 1
```

Which is better than? `r emo::ji('confused')`
---
name: vectorization_benchmark

## Repeating actions &mdash; vectorization

Let us compare the time of execution of the vectorized version (vector with 10,000 elements):
```{r for.loop.avoid.timing, echo=F}
vec <- c(1:1e6)
ptm <- proc.time()
vec <- vec + 1
proc.time() - ptm # vectorized
```

--

to the loop version:
```{r for.loop.avoid.timing2, echo=F}
vec <- c(1:1e6)
ptm <- proc.time()
for (i in vec) {
  vec[i] <- vec[i] + 1
}
proc.time() - ptm # for-loop
```

---
name: base_vectorize
## Vectorization &mdash; the problem
```{r is_a_droid}
is_a_droid <- function(x) {
  droids <- c('2-1B', '4-LOM', '8D8', '0-0-0', 'AP-5', 'AZI-3', 'Mister Bones', 'BB-8', 'BB-9E', 'BD-1', 'BT-1', 'C1-10P', 'C-3PO', 'R2-D2')
  if (x %in% droids) {
    return(T)
  } else {
    return(F)
  }
}

test <- c('Anakin', 'Vader', 'R2-D2', 'AZI-3', 'Luke')
is_a_droid(test)
```

---
name: base_vectorize2
## Vectorization &mdash; the solution(s)

The `base::Vectorize` way:
```{r vec_is_a_droid}
vectorized_is_a_droid <- base::Vectorize(is_a_droid, vectorize.args = c('x'))
vectorized_is_a_droid(test)
```

--

The `apply*` way:
```{r, error=T}
apply(as.matrix(test), FUN = is_a_droid, MARGIN = 1)
```

--

```{r}
lapply(test, FUN=is_a_droid) %>% unlist()
```

--

```{r}
sapply(test, is_a_droid)
```

---
name: base_vectorize3
## Vectorization &mdash; the solution(s)

The `vapply`:
```{r}
vapply(test, is_a_droid, FUN.VALUE = TRUE)
```

--

```{r}
vapply(test, is_a_droid, FUN.VALUE = 1)
```

--

```{r, error = T}
vapply(test, is_a_droid, FUN.VALUE = c(1,0))
vapply(test, is_a_droid, FUN.VALUE = 'a')
```

--

Or the `purrr` way:
```{r}
purrr::map(test, is_a_droid) %>% unlist()
```

---
name: recursion
## Recursion
When we explicitely repeat an action using a loop, we talk about **iteration**. We can also repeat actions by means of **recursion**, i.e. when a function calls itself. Let us implement a factorial $!$: 
```{r rec.fact, echo=T}
factorial.rec <- function(x) {
  if (x == 0 || x == 1)
    return(1)
  else
    return(x * factorial.rec(x - 1)) # Recursive call!
}
factorial.rec(5)
```

---
name: rec_eq_iter
## Recursion = iteration?
Yes, every iteration can be converted to recursion (Church-Turing conjecture) and *vice versa*. It is not always obvious, but theoretically it is doable. Let's see how to implement *factorial* in iterative manner:

```{r rec.fact.iter, echo=T}
factorial.iter <- function(x) {
  if (x == 0 || x == 1)
    return(1)
  else {
    tmp <- 1
    for (i in 2:x) {
      tmp <- tmp * i
    }
    return(tmp)  
  }
}
factorial.iter(5)
```

---
name: rec_eq_iter_really
## Recursion == iteration, really?

More writing for the iterative version, right? What about the time efficiency?  
The recursive version:
```{r rec.fact.timing, echo=F}
peakRAM::peakRAM(factorial.rec(2000))
```
And the iterative one:
```{r iter.fact.timing, echo=F}
peakRAM::peakRAM(factorial.iter(2000))
```

---
name: loops_avoid_growing
## Loops &mdash; avoid growing data
Avoid changing dimensions of an object inside the loop:
```{r avoid.growing, echo=T}
v <- c() # Initialize
for (i in 1:100) {
  v <- c(v, i)
}
```

--

It is much better to do it like this:
```{r avoid.growing2, echo=T}
v <- rep(NA, 100) # Initialize with length
for (i in 1:100) {
  v[i] <- i
}
```

--


<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end-slide
class: end-slide, middle
count: false

# Thank you. Questions?

```{r,echo=FALSE,child="assets/footer-presentation.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
rmarkdown::render("presentation.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("presentation.html",output="presentation.pdf")
```
